Open-source landscape for 3D CSEM modelling

### 2020-04-05 manuscript/Manuscript.tex

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Write 1-2 sentence(s) each:
%
\begin{enumerate}
  \item Principal OBJECTIVES AND SCOPE OF THE WORK
  \item Methodology
  \item Results
  \item Conclusions
\end{enumerate}
%
\end{abstract}

@@

Publications regarding 3D modelling in electromagnetic methods started to
appear as early as the 70's and 80's. These early publications where all
integral equation (IE) methods, having an anomaly embedded within a layered
medium, for loop-loop type transient EM measurements \citep{GJI.74.Raiche,
GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and magnetotelluric (MT)
measurements \citep{GEO.84.Wannamaker}.

In the 90's computer became sufficiently powerful that 3D modelling gained
traction, which resulted amongst other in the publication of the book
\emph{Three-Dimensional Electromagnetics}, \cite{B.SEG.99.Oristaglio} by the
SEG. Often cited publications from that time are \cite{RSC.94.Mackie} (3D MT
calculation), \cite{RS.94.Druskin} (frequency- and time-domain modelling using
a Yee grid and a global Krylov subspace approximation), and
\cite{RS.96.Alumbaugh, GJI.97.Newman} (low-to-high frequency calculation on
massively parallel computers).

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of publications. The amount
of available numerical solutions can be overwhelming and is part of the reason
why there are hundreds of publications about the topic. There are the different
methods to solve Maxwell's equation, such as the IE method
\citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the DE
method, for instance FD \citep{IEEE.66.Yee, RSC.94.Mackie, RS.94.Druskin,
GEO.09.Streich, CAG.13.Sommer}, FE \citep{GJI.11.Schwarzbach, GEO.12.daSilva,
GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang}, FV \citep{EM.90.Madsen,
ECP.07.Haber, GEO.14.Jahandari}, and FI \citep{PIER.01.Clemens, GP.06.Mulder}.
And these are just the most common ones.

@@

A very well written overview up to the year 2005 of the different approaches to
3D EM modelling is given by \cite{SG.05.Avdeev}. In the last 15 years the
publications with regards to 3D EM  modelling grew tremendously, driven
probably by improved computer powers.

@@

\paragraph{Overviews: From \cite{SG.05.Avdeev}}

Over the last decade, the EM induction community had three large international
meetings entirely devoted to the theory and practice of three-dimensional (3-D)
electromagnetic (EM) modelling and inversion (see Oristaglio and Spies, 1999;
Zhdanov and Wannamaker, 2002; Macnae and Liu, 2003). In addition, two special
issues of Inverse Problems dedicated to the same subject have been recently
published (Lesselier and Habashy, 2000; Lesselier and Chew, 2004).

«\emph{I would like to conclude this review with the following general remark.
  The most important challenge that faces the EM community today is to convince
  software developers to put their 3-D EM forward and inverse solutions into
  the public domain, at least after some time. This would have a strong impact
  on the whole subject and the developers would benefit from feedback regarding
the real needs of the end-users.}»

=> Use this quotation as a hook.

\textbf{Just a list of ideas and notes so far.}

List codes, particularly papers presented that are, potentially, available or
not, but only by contacting the author. Mention change in coding practices,
development in the open, collaboration, bug reports, better documentation and
testing, etc. Example: \cite{CAG.17.Cai} (open or not?)

@@

Common features of our codes is:

\begin{itemize}
  \item Not just open source, but also developed in the open and increasingly
    building a global community - also mention other projects as, e.g., pyGIMLi
    \citep{CAG.17.Rucker} and Fatiando (e.g., \cite{JOSS.18.Uieda}).
  \item All codes we compare work in the Python eco-system (hard work
    underlying the Python API can differ though).
  \item Also modern dev, which includes CI testing, good documentation, issue
    tracking, and extensive discussion through Slack, Gitter, or similar, with
    dozens to hundreds of members (SimPEG 09-09-2019: 189 members), which
    constitutes a vibrant community and combines a huge spectrum from coders to
    first time users, and from academia to industry.
\end{itemize}

@@

\begin{itemize}
  \item Mention all use cases (ground water, geothermal, oil and gas, civil
        engineering, hazards).
  \item Mention other fields.
  \item Mention MT comparison paper.
\end{itemize}

Many 3D codes exist since long (SEG 3D book by oristaglio, etc). However, there
are a few key differences recently:

\begin{itemize}
  \item Codes are \emph{just available} (you don't have to write someone and
    ask if you could have the code)
  \item Codes are \emph{easy installable} (e.g., ``pip install emg3d``)
  \item Codes are well documented and come with a lot of examples
  \item It is easy to get involved (GitHub/GitLab) and contribute
  \item It is easy to get help (Slack, GitHub/GitLab)
\end{itemize}

@@

time domain and frequency domain

1D, 2D, we look at 3D.

@@

@@

Currently \emg3d can be used to model the electric and magnetic fields due to
arbitrary finite length electric dipole sources for a single frequency at a
time. Helper functions to calculate time-domain responses are in the works
(should be ready by publishing this article). Current development is focused
on adding inversion capabilities; averaging routines to implement strong
topography; and general anisotropy.

The 3D solver \emg3d can be found in the GitHub organization of \empymod
(`empymod.github.io <https://empymod.github.io>`_). The modeller \empymod
\citep{GEO.17.Werthmuller} is a 1D code which can calculate electric or
magnetic responses due to a 3D electric or magnetic source in a layered-earth
model with vertical transverse isotropic (VTI) resistivity, VTI electric
permittivity, and VTI magnetic permeability, from very low frequencies (DC) to
very high frequencies (GPR).

@@

We would like to thank Paulo Menezes for the help and explanations with regards
to the Marlim R3D model and corresponding CSEM data, and for making their
actual calculation model available under an open-source license.

>>>>

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Write at the end; one or two sentence(s) each:
%
\begin{enumerate}
  \item Principal OBJECTIVES AND SCOPE OF THE WORK
  \item Methodology
  \item Results
  \item Conclusions
\end{enumerate}
%
\end{abstract}

@@

Controlled-source electromagnetic (CSEM) measurements is a frequently applied
method in exploration geophysics for various applications, such as ground
water, geothermal, oil and gas, mining, civil engineering, or geo-hazards.
Modelling these electromagnetic fields is therefore of great interest in order
to understand the measured data. Publications regarding 3D modelling in
electromagnetic methods started to appear as early as the 1970's and 1980's.
These early publications where all integral equation (IE) methods, having an
anomaly embedded within a layered medium, for loop-loop type transient EM
measurements \citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman}
and magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

In the 1990's computer became sufficiently powerful that 3D modelling gained
traction, which resulted amongst other in the publication of the book
\emph{Three-Dimensional Electromagnetics} \citep{B.SEG.99.Oristaglio} by the
SEG. Often cited publications from that time are \cite{RSC.94.Mackie}, 3D MT
computation; \cite{RS.94.Druskin}, frequency- and time-domain modelling using a
Yee grid and a global Krylov subspace approximation; and \cite{RS.96.Alumbaugh,
GJI.97.Newman}, low-to-high frequency computation on massively parallel
computers.

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of publications. The amount
of available numerical solutions can be overwhelming and is part of the reason
why there are hundreds of publications about the topic. There are the different
methods to solve Maxwell's equation, such as the IE method
\citep{GJI.74.Raiche, RS.02.Hursan, GEO.06.Zhdanov, GP.10.Tehrani,
CAG.16.Kruglyakov, MGS.17.Kruglyakov} and different variations of the
differential equation (DE) method, for instance finite differences (FD)
\citep{IEEE.66.Yee, RSC.94.Mackie, RS.94.Druskin, GEO.09.Streich,
CAG.13.Sommer}, finite elementes (FE) \citep{GJI.11.Schwarzbach,
GEO.12.daSilva, GJI.13.Puzyrev, GJI.13.Grayver, SEG.16.Zhang}, finite volume
(FV) \citep{EM.90.Madsen, ECP.07.Haber, GEO.14.Jahandari}, and finite
integration technique (FIT) \citep{PIER.01.Clemens, GP.06.Mulder}. And these
are just the most common ones.

@@

A very well written overview up to the year 2005 of the different approaches to
3D EM modelling is given by \cite{SG.05.Avdeev}. In the last 15 years the
publications with regards to 3D EM  modelling grew tremendously, driven
probably by improved computer powers. Avdeev finished his review with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}»

It is this topic that we want to address, open-source 3D CSEM codes. Some big
exploration companies and service companies have their own codes, as well as
certain research consortia. But how about codes that are open-source?
Traditionally research code was often available upon request by the author.
Codes distributed that way often come with the request to not share the code,
or they often come without a license attached at all (which means that they are
not open-source). Also, at times there can be a significant hurdle to install
a code.

However, the meaning of open-source evolved rapidly in the last decades. Today
open-source generally not only means that a code is available (with a proper
license). It includes much more today, such as that the code is hosted online,
is under version control, has the possibility to file issues and make pull
requests to fix bugs. Good codes also have continuous integration that includes
extensive testing of the code base and online hosted documentation that
includes the API. As such the term has shifted from purely open-source code
to development in the open and increasingly building a global community. There
are a number of projects within the realm of geophysical exploration, besides
the ones we present here, e.g. pyGIMLi \citep{CAG.17.Rucker} and Fatiando
\citep{JOSS.18.Uieda}.

Newer \todo{wrong term, think about it}programming environments  have also
brought much simplifications in the installation process and have most
importantly simplified it for code developers to make their codes easily
available and installable. All our presented codes are in the Python
environment, and each code can be install with a single command (which includes
downloading and installation).

@@

@@

@@

@@

1D, 2D, we look at 3D.
time domain and frequency domain

@@

All the codes presented here
\begin{itemize}
  \item are \emph{just available} (you don't have to write someone and ask if
    you could have the code);
  \item are \emph{easy installable} (through \texttt{pip} or \texttt{conda});
  \item are under \emph{version control} and online either on GitHub or on
    GitLab, which includes possibilities of easy involvement by either creating
    an issue, reporting a bug, or filing a pull request; and
  \item are well documented and come with a lot of examples.
\end{itemize}

\begin{itemize}
  \item Code (license): \emg3d (Apache-2.0), custEM (LPGL-3.0), PETGEM
    (GPL-3.0), and \simpeg (MIT).
  \item Hosting: GitHub (\emg3d, \petgem, \simpeg); GitLab (\custem)
\end{itemize}

@@

Currently \emg3d can be used to model the electric and magnetic fields due to
arbitrary finite length electric dipole sources for a single frequency at a
time. Helper functions to compute time-domain responses are in the works
(should be ready by publishing this article). Current development is focused
on adding inversion capabilities; averaging routines to implement strong
topography; and general anisotropy.

The 3D solver \emg3d can be found in the GitHub organization of \empymod
(`empymod.github.io <https://empymod.github.io>`_). The modeller \empymod
\citep{GEO.17.Werthmuller} is a 1D code which can compute electric or
magnetic responses due to a 3D electric or magnetic source in a layered-earth
model with vertical transverse isotropic (VTI) resistivity, VTI electric
permittivity, and VTI magnetic permeability, from very low frequencies (DC) to
very high frequencies (GPR).

@@

We would like to thank Paulo Menezes for the help and explanations with regards
to the Marlim R3D model and corresponding CSEM data, and for making their
actual computation model available under an open-source license.
